---
title: "Class 10 part 2"
author: "Neva Olliffe PID A69026930"
format: html
---


# Finish lab 10 Part 4: Adenylate Kinase
We will use the `bio3d` package for this analysis that starts with a single sequence.

We will also use the `msa` package from BioConductor.
First we need to install the `BiocManager` package. First, we install `BiocManager` with `install.packages()`. 

We use `BiocManager::install()` to install any other BioConductor packages. 


## Q10. 

```{r}
library(bio3d)
aa <- get.seq("1ake_A")

```


```{r}
aa
```
## Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 
214 amino acids long.

```{r}
# Blast or hmmer search 
#b <- blast.pdb(aa)
```
```{r}
# Plot a summary of search results
#hits <- plot(b)
```
```{r}
# List out some 'top hits'
#head(hits$pdb.id)
```
> Let's save our results.

```{r}
save(hits, b, file="blast_results.Rds")
```

# Test
```{r}
load("blast_results.Rds")
hits
```



```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
```{r}
# Vector containing PDB codes for figure axis
ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
plot(pdbs, labels=ids)
```

```{r}
# Annotate pdb sequences 
anno <- pdb.annotate(ids)
unique(anno$source)

```

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)

```

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```







