---
title: "Class 5: Data Visualization with ggplot"
author: "Neva Olliffe (A69026930)"
format: pdf
---

# Using GGPlot

##Basics with cars
Load ggplot2 package before use. 
```{r}
library(ggplot2)

ggplot(cars)
```
```{r}
head(cars)
```
We can use the base R graphics system to plot. 
```{r}
plot(cars)
```

ggplot requires more information than the base `plot()` function:
- Data, in a `data.frame`
- aes - how the data map to the plot 
- geoms - how things are drawn

```{r}
ggplot(cars)+
  aes(x = speed, y = dist) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Speed and stopping distance", 
       x = "Speed (mph)", 
       y = "Distance (feet)", 
       caption = "Dataset: cars") + 
  theme_bw()
```

## Switching to gene expression dataset
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
Exploring the dataset
```{r}
nrow(genes)
colnames(genes)
ncol(genes)
table(genes$State)

```

Plot the dataset
```{r}
p <- ggplot(genes) + 
  aes(x = Condition1, y = Condition2, col=State)+ 
  geom_point() + 
  scale_color_manual(values=c("blue", "gray", "red")) + 
  labs(title = "Gene Expression Changes Upon Drug Treatment", 
       x = "Control (no drug)", 
       y = "Drug Treatment")
p
```

## Switch to  gapminder dataset
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

### Add dplyr and filter for the year 2007
```{r}
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```

Simple scatterplot for 2007
```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, col = continent, size = pop) + 
  geom_point(alpha = 0.5) +
  scale_size_area(max_size = 10)
```

###Recapitulate the above process for 1957
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, col = continent, size = pop) + 
  geom_point(alpha = 0.7) + 
  scale_size_area(max_size = 15)

```

### Graph both 2007 and 1957
```{r}
gapminder_1957_2007 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957_2007) + 
  aes(x = gdpPercap, y = lifeExp, col = continent, size = pop) + 
  geom_point(alpha = 0.7) + 
  scale_size_area(max_size = 10) + 
  facet_wrap(~year)
```

## Bar charts
Filter for the 5 biggest countries
```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>%
  arrange(desc(pop)) %>%
  top_n(5, pop)

gapminder_top5
```


```{r}
ggplot(gapminder_top5) + 
  geom_col() + 
  aes(x = country, y = pop)
```

```{r}
ggplot(gapminder_top5) + 
  aes(x = country, y = lifeExp, fill = continent) + 
  geom_col()
```


```{r}
ggplot(gapminder_top5) + 
  aes(x = country, y = lifeExp, fill = lifeExp) + 
  geom_col()
```

```{r}
ggplot(gapminder_top5) + 
  aes(x = reorder(country, -pop), y= pop, fill = country) + 
  geom_col(col="black")
```

### Flipping bar charts
```{r}
USArrests$State <-rownames(USArrests)

ggplot(USArrests) + 
  aes(x = reorder(State, Murder), y = Murder) + 
  geom_point() + 
  geom_segment(aes(x = State, 
                   xend = State, 
                   y = 0, 
                   yend = Murder), 
               color = "blue") + 
  coord_flip()
```




```{r}
ggplot(mtcars, aes(mpg, disp)) + geom_point()
```
```{r}
ggplot(mtcars) + aes(x=mpg, y=disp) + geom_point()
```

