---
author: "Neva Olliffe PID A69026930"
title: "Class08 halloween"
format: pdf
---


# 1. Import the data
```{r}
candy_url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy = read.csv(url(candy_url), row.names=1)
head(candy)

```

## Q1. How many different candy types are in this dataset?
```{r}
dim(candy)
```
85 types of candy.

## Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
```
38 types of candy are fruity. 

#2. What's your favorite candy?
```{r}
candy["Twix",]$winpercent
candy["Kit Kat",]$winpercent
candy["Tootsie Roll Snack Bars",]$winpercent
```
Twix wins 81.6% of the time. 

```{r}
#library("skimr")
skimr::skim(candy)
```
## Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
Win percent looks like a very different scale. 

## Q7. What do you think a zero and one represent for the candy$chocolate column?
A zero stands in for `False` and a 1 for `True`. 

## Q8. Plot a histogram of winpercent values
```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) + 
  geom_histogram(binwidth = 5)
```
## Q9. Is the distribution of winpercent values symmetrical?
It is slightly skewed right. 

## Q10. Is the center of the distribution above or below 50%?
```{r}
median(candy$winpercent)
mean(candy$winpercent)
```
The center of the distribution is just slightly below 50% if you use the median, or just above if you use the mean. 

## Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
choc_candy <- candy[as.logical(candy$chocolate), "winpercent"]
fruit_candy <- candy[as.logical(candy$fruity), "winpercent"]

mean(choc_candy)
mean(fruit_candy)

```
Chocolate candy is ranked higher on average than fruit candy.

## Q12. Is this difference statistically significant?
```{r}
t.test(choc_candy, fruit_candy)
```
# 3. Overall candy rankings
## Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```
## Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) + 
  geom_col()
```

## Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) + 
  geom_col()
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) + 
  theme(text = element_text(size=6))

```
## Q17. What is the worst ranked chocolate candy?
Sixlets.

## Q18. What is the best ranked fruity candy?
Starburst.

# Taking a look at pricepoint. 
```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols)  
  #+ geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
## Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
```{r}
row.names(candy)[which.max(candy$winpercent/candy$pricepercent)]
```
## Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )

```
Nik L Nip is the most expensive candy and the least popular of the top 5 most expensive candies.

## Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().
```{r}
priced_candy <- reorder(rownames(candy), candy$pricepercent)

ggplot(candy, aes(x = pricepercent, priced_candy)) + 
  geom_point() + 
  geom_segment(aes(x = pricepercent, xend = 0, yend = priced_candy, 
                   y = priced_candy)) + 
    theme(text = element_text(size=6))

```

# 5 Exploring the correlation structure
```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```
## Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Fruity & chocolate. 

## Q23. Similarly, what two variables are most positively correlated?
chocolate & winpercent. 

# 6. Principal Component Analysis
```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)

```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

#p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
 # theme(legend.position = "none") +
  #labs(title="Halloween Candy PCA Space",
    #   subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       #caption="Data from 538")

```

```{r}
#library(plotly)
#ggplotly(p)

```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
## Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Fruity, hard, and pluribus are picked up strongly. This makes sense because many fruity candies come in multiple packs (skittles, starbursts, etc) which is less common for chocolate candies. 




